Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-14 aistats.tex
small things

The density of an exponential family for data $x \in \cX$ reads
\begin{equation}
	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
	 \label{eq:def_expfamily}
\end{equation}
where  $\nat$ is called natural (or primal) parameter.
It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) a base measure$\nu$ on $\cX$.
Since the exponential is positive, $p$ has the same support as $\nu$.
The function $\logpart$ is called the log-partition function that acts as a normalization term, since
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
\end{align}

This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

For convenience, in this paper we focus on steep, regular exponential families with minimal statistic $T$.
Then $\logpart$ is a strictly convex function of Legendre type,
and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex
\citep{barndoffnielsen2014information}. 
Also, when explicit, we write the random variable $T = T(X)$.

@@

We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the log-partition function
\begin{align}
	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
\end{align}
which matches the common notion of \textit{entropy} in information theory.
By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}). More concisely,
\aligns{
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
}

\begin{figure}[t] 
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}

\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
We now introduce the Bregman divergence induced by $\logpart$, that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
    \label{eq:defBregman}
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. In general, bregman distances are non-symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$. Finally, it is equal to the divergence of its convex conjugate with switched arguments,
\begin{align}
	\bregman (\nat ; \nat^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
    \label{eq:bregman_switch}
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
    \label{eq:def_prior}
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is the number of fictive data points observed from a distribution with natural parameter $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).} 
Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD|\nat)p(\nat)$.
Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
\begin{align}
	p(\nat \cond \mD_n)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
which reaches its maximum in $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)~ \eqref{eq:defMLE}.
In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.

\paragraph{Optimization}
We can rewrite~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$. 
Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
\nabla\conj(\hat \nat_{n}) 
= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
}
which we recognize as the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ and step-size schedule $\lr_n$ initialized at $\nat_0$.
\emph{MAP is an instantiation of SMD.}
Note that this analogy does not hold for the MLE %($n_0=0$)
as it does not have any initialization to start from.

\section{Open Problem}
Let us assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
As we will see, this is already an interesting setting.
We wish to estimate how well the MLE or a MAP model the true distribution.

A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
Using~\eqref{eq:defNLL} and~\eqref{eq:bregman_switch}, we can also write the KL with Bregman divergences
\alignn{
	\KL(p_{\nat^*} || p_\nat)
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
}

How does this quantity behave when $\nat$ is the MLE or a MAP ? 
For instance can we get upper bounds on the expected KL ?
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
\end{align}
where the outer expectation is on the dataset $\mD$.
More explicitly, can we write an upper bound that does not involve this expectation over the dataset?
Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators, when the statistical loss is given by the KL divergence ?

Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights}.
There are also results for special cases, such as $\cN(0, \sigma^2)$ (See \S\ref{sec:example}). 
But we are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n \lesssim d$.

{\bf Unboundedness of~\eqref{eq:bregmanMLE}.}
While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, 
for instance in Gaussian covariance estimation when $n \leq d + 1$.
Even worse, with categorical variables, there is always a non-zero probability to never sample one of the categories. In such the MLE will give zero weight to this category and $KL(p_{\nat^*} || p_\text{MLE} ) = +\infty$. 
This means that the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality but this is a difficult endeavor.
These examples are rather making a case for regularized estimators such as MAP, for which we may find interesting upper bounds.

@@

\subsection{Quadratic Case}
As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $I$ and unknown mean 
\[ 
	\cX=\real^d,\quad \nu(dx) = \exp({\textstyle \half[-\|x\|^2]})dx,\quad T(x)=x.
\]
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
	=\frac{\Var(T)}{2n}
	=\frac{1}{2n}
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields to
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

>>>>

The density of an exponential family for a sample $x$ is
\begin{equation}
	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
	 \label{eq:def_expfamily}
\end{equation}
where  $\nat$ is called natural (or primal) parameter.
It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, 
and 2) a base measure $\nu$ on $\cX$.
Since the exponential is positive, $p$ has the same support as $\nu$.
The log-partition function $\logpart$ acts as a normalization term, since
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \;.
\end{align}
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

For convenience, we focus on steep, regular exponential families with minimal statistic $T$.
Then $\logpart$ is a strictly convex function of Legendre type,
and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex
\citep{barndoffnielsen2014information}. 
When explicit, we write the random variable $T = T(X)$.

@@

We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
\begin{align}
	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
\end{align}
which is the common notion of \textit{entropy} in information theory.
By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$ 
(see Fig.~\ref{fig:duality}), giving
\aligns{
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
}
\begin{figure}[t] 
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}%
%
%\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
% We now introduce the Bregman divergence induced by $\logpart$, 
% that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
The Bregman divergence induced by $\logpart$, 
measures the discrepancy between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
    \label{eq:defBregman}
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. 
In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$. 
Finally, it is equal to the divergence of its convex conjugate with switched arguments,
\begin{align}
	\bregman (\nat ; \nat_0)
    = \bregmanconj ( \meanp_0 ; \meanp) \; .
    \label{eq:bregman_switch}
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
    \label{eq:def_prior}
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is an exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is the number of fictive data points observed from a distribution with natural parameter $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).} 
Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD|\nat)p(\nat)$.
Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
\begin{align}
	p(\nat \cond \mD_n)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
which reaches its maximum attempts $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- we have no samples from the prior -- we recover the MLE~\eqref{eq:defMLE}.
We write $\MAPt$ for the MAP, and view the MLE as a special case.

\paragraph{Optimization}
We can rewrite~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$. 
Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n}) 
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
}
which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ 
and step-size schedule $\lr_n$, initialized at $\nat_0$.
\emph{MAP is an instantiation of SMD.}
Note that this analogy does not hold for the MLE %($n_0=0$)
as it does not have any initialization to start from.
\fdk{It does, you just forget the initial condition after the first step with step-size 1?}

\section{Open Problem}
% Assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
% We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
% As we will see, this is already an interesting setting.
% We wish to estimate how well the MLE or a MAP model the true distribution.

Assume we observe a dataset $\mD$ drawn i.i.d. from an exponential family distribution 
with parameters $\nat^*$, $p(\cdot \cond\nat^*)$.
We wish to quantify how well the MLE or a MAP models the true distribution.

A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
Using~\eqref{eq:defNLL} and~\eqref{eq:bregman_switch}, we can also write the KL with Bregman divergences
\alignn{
	\KL(p_{\nat^*} || p_\nat)
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
}
How does this quantity behave when $\nat$ is the MLE or MAP?
Can we get upper bounds on the expected KL?
\begin{align}
	\label{eq:bregmanMLE}
	\text{MLE: } &\quad \expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
	\label{eq:bregmanMAP}
	\text{MAP: } &\quad \expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
\end{align}
where the expectation is on the data $\mD$.
More explicitly, is there an upper bound that does not involve this expectation over the dataset?
Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators 
when the loss is the KL divergence?

Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights},
or results for special cases such as $\cN(0, \sigma^2)$ (see \S\ref{sec:example}). 
But we are yet to find a solution encompassing a broad range of exponential families, 
and applicable to small sample sizes $n \lesssim d$.

{\bf Unboundedness of~\eqref{eq:bregmanMLE}.}
While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, 
for instance when estimating the covariance of a Gaussian when $n \leq d + 1$.
Even worse, with categorical variables, there is a non-zero probability to never sample one of the categories. 
In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$,
and the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
Instead of taking the expectation, one might want to bound the risk in high probability, 
without resorting to Markov inequality but this is a difficult endeavor.
These examples make a case for regularized estimators such as MAP, 
for which we may find upper bounds.

@@

\subsection{Quadratic Case}
As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with known variance $I$,
\[ 
	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
\]
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
	=\frac{\Var(T)}{2n}
	=\frac{1}{2n}
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields to
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.
